---
title: "Simulating Trees (and Trait Data)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating Trees}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.5,
  dpi = 300,
  out.width = "100%",
  eval = requireNamespace("ggtree", quietly = TRUE) && requireNamespace("ggplot2", quietly = TRUE)
)
```

```{r setup, message=FALSE}
library(scTreeSim)
library(ggtree)
library(dplyr)
source("helpers.R")  # plotting helpers of the vignettes: plot_tree() and a colour-blind safe palette
```

This vignette covers the ways to simulate a tree (by origin time or by number of taxa) under the age-dependent branching process.
Sampling and pruning effects are shown, single- and multi-type trees are presented.
The gene expression and a binary trait on top of a single fixed tree is simulated. 
Barcodes have their own vignette, `vignette("simulate_barcodes")`.

All trees are `treedata` objects (from `treeio`) and are drawn with `ggtree`, with time before present on the axis: 
the dashed lines mark the **origin** of the process and the **present**.

# Simulating trees

Trees are generated by an age-dependent branching process: the lifetime of a cell is
gamma-distributed (`shape`, `scale`; its mean is `shape * scale`, or give `mean_lifetime`),
and at the end of its life a cell dies (with probability `death_prob`) or divides in two.

```{r}
shape <- 100
av_lifetime <- 1
d <- 0.1
scale <- av_lifetime / shape
```

## By origin time

Simulates forward from a single ancestor for a fixed amount of time (`origin_time`), 
so the number of resulting tips is random. The tree contains the lineages that died
along the way (grey tips); the tips alive at the present are green.

```{r}
t_or <- 6
set.seed(5)
tree_by_time <- sim_adb_origin_complete_fast(t_or, scale, shape, d)
plot_tree(tree_by_time, color = "status")
```

## By number of taxa

Simulates forward until a fixed number of living taxa (`ntaxa`) is reached,
so the time span of the tree is random instead (compare the origin on the axis).

```{r}
ntaxa <- 2^5
set.seed(5)
tree_by_ntaxa <- sim_adb_ntaxa_complete_fast(ntaxa, scale, shape, d)
plot_tree(tree_by_ntaxa, color = "status")
```

## Sampling and pruning

`sim_adb_origin_samp()` (and its `ntaxa` counterpart `sim_adb_ntaxa_samp()`) additionally samples
tips at present with probability `sampling_prob` and prunes extinct and unsampled
lineages, giving the tree that is actually observed.

```{r}
set.seed(5)
tree_sampled <- sim_adb_origin_samp(t_or, scale, shape, d, sampling_prob = 0.2)
plot_tree(tree_sampled, color = "status")
```

By default (`collapse = TRUE`) the divisions that lost one of their two lineages are removed, so every
internal node has two children. With `collapse = FALSE` these nodes are kept as
single-child nodes: the same tree has more nodes, but no information on the types and divisions
is lost, as required for the multi-type models and the division-triggered spikes of the binary trait model below.
With the same seed, the simulated tree and the sampled tips are identical in both cases:

```{r}
set.seed(5)
tree_kept <- sim_adb_origin_samp(t_or, scale, shape, d, sampling_prob = 0.2, collapse = FALSE)
c(collapsed = nrow(tree_sampled@data), kept = nrow(tree_kept@data))

plot_tree(tree_kept, color = "status", show_nodes = T)
```

# Multi-type trees

Passing per-type parameters and type-transition matrices (`asym_trans_prob`, `sym_trans_prob`)
simulates a tree in which cells switch between discrete types: at a division, an
asymmetric transition makes one daughter change type, a symmetric one lets both
daughters take the type of their parent. The tree starts in `origin_type`, and the root edge is
colored accordingly. This tree is used for the data below and in `vignette("simulate_barcodes")`.

```{r}
ntype <- 2
origin_type <- 0
shape <- c(100, 1)
av_lifetime <- rep(1, ntype)
scale <- av_lifetime / shape
rho <- 0.8
t_or <- 7
d <- c(0.1, 0.1)

r <- 0.11
Xi_as <- matrix(c(0.0, 0.0, r, 0.0), nrow = ntype, ncol = ntype)
Xi_s  <- matrix(c(1 - r, 0.0, 0.0, 1.0), nrow = ntype, ncol = ntype)

# collapse = FALSE keeps every division node in the tree, including ones
# with only one surviving child -- required for the division-triggered
# spikes used in the binary trait model below
set.seed(2)
mt_tree <- sim_adb_origin_samp(t_or, scale, shape, d, rho, origin_type,
                               asym_trans_prob = Xi_as, sym_trans_prob = Xi_s, collapse = FALSE)
# a simulation with too few tips returns NULL: try another seed
stopifnot(!is.null(mt_tree))
```

```{r, fig.height = 6}
plot_tree(mt_tree, color = "type", show_nodes = T)
```

# Gene expression (Ornstein-Uhlenbeck)

Simulates one continuous trait per node under an OU process with a
type-dependent optimum (`theta`). `alpha = 0` reduces this to Brownian
motion.

```{r}
# `type` of theta has the same class as `type` in the tree (integer)
theta <- data.frame(type = c(0L, 1L), theta = c(0, 3))
stopifnot(all(unique(mt_tree@data$type) %in% theta$type))

expr <- sim_expression_ou(mt_tree, alpha = 0.5, sigmasq = 0.2, root_value = 0, theta = theta)
head(expr)
```

```{r}
expr %>%
  left_join(mt_tree@data %>% select(node, status), by = "node") %>%
  filter(status == 1) %>%
  ggplot2::ggplot(ggplot2::aes(x = factor(type), y = expr, fill = factor(type))) +
  ggplot2::geom_boxplot(show.legend = FALSE) +
  ggplot2::scale_fill_manual(values = type_cols) +
  ggplot2::labs(x = "type", y = "expression at the tips")
```

# Binary trait, with error and division-triggered spikes

A binary trait evolving under a 2-state CTMC, with optional
gamma-distributed mutation spikes at branches where a division occurred,
and optional false positive/negative error at the tips.

```{r}
bc_binary <- sim_binary_ontree(mt_tree, l = 10, gamma_spike = TRUE,
                               spike_mean = 0.3, spike_shape = 2,
                               alpha = 0.05, beta = 0.02)
head(bc_binary)
```
