---
title: "Basic Tree Simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic Tree Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(scTreeSim)
library(ggtree)
library(dplyr)
set.seed(12)
```

This vignette covers the two ways to simulate a tree (by origin time or by
number of taxa), how to simulate a multi-type tree, and how to layer each of
the package's data modalities : gene expression, sequential and
non-sequential barcodes, and a binary trait on top of a single fixed tree.

# Simulating trees

## By origin time

Simulates forward from a single ancestor for a fixed amount of time
(`t_or`), so the number of resulting tips is random.

```{r}
shape <- 100
av_lifetime <- 1
d <- 0.08
scale <- av_lifetime / shape
t_or <- 6

tree_by_time <- sim_adb_origin_complete_fast(t_or, scale, shape, d)
plot(tree_by_time@phylo, show.tip.label = FALSE, root.edge = TRUE)
```

## By number of taxa

Simulates forward until a fixed number of extant taxa (`ntaxa`) is reached,
so the tree's time span is random instead.

```{r}
ntaxa <- 2^6

tree_by_ntaxa <- sim_adb_ntaxa_complete_fast(ntaxa, scale, shape, d)
plot(tree_by_ntaxa@phylo, show.tip.label = FALSE, root.edge = TRUE)
```

Both return a `treedata` object (from `treeio`). Extinct lineages are still
present in these trees; sampling at present and pruning extinct lineages is
covered next.

## Sampling and pruning

`sim_adb_origin_samp()` (and its `ntaxa` counterpart) additionally samples
tips at present with probability `rho` and prunes extinct/unsampled
lineages, giving the tree that is actually observed.

```{r, fig.show="hold", out.width="40%"}
rho <- 0.5

dat_sampled <- sim_adb_origin_samp(t_or, scale, shape, d, rho)
plot(dat_sampled@phylo, show.tip.label = FALSE, root.edge = FALSE)
```

# Multi-type trees

Passing per-type rates and type-transition matrices (`Xi_as`, `Xi_s`)
simulates a tree in which cells switch between discrete types. This is the
tree used for every data modality in the rest of this vignette.

```{r}
ntype <- 2
origin_type <- 0
shape <- c(100, 1)
av_lifetime <- rep(1, ntype)
scale <- av_lifetime / shape
rho <- 0.8
t_or <- 7
d <- c(0.1, 0.1)

r <- 0.11
Xi_as <- matrix(c(0.0, 0.0, r, 0.0), nrow = ntype, ncol = ntype)
Xi_s  <- matrix(c(1 - r, 0.0, 0.0, 1.0), nrow = ntype, ncol = ntype)

# collapse = FALSE keeps every division node in the tree, including ones
# with only one surviving child -- required for the division-triggered
# spikes used in the binary trait model below
dat2 <- sim_adb_origin_samp(t_or, scale, shape, d, rho, origin_type,
                             Xi_as = Xi_as, Xi_s = Xi_s, collapse = FALSE)

ggtree(dat2, aes(color = type)) +
  geom_rootedge(color = "#F8766D") +
  geom_tippoint()
```

# Gene expression (Ornstein-Uhlenbeck)

Simulates one continuous trait per node under an OU process with a
type-dependent optimum (`theta`). `alpha = 0` reduces this to Brownian
motion.

```{r}
theta <- data.frame(type = factor(c(0, 1)), theta = c(0, 3))
stopifnot(all(levels(dat2@data$type) %in% levels(theta$type)))

expr <- sim_expression_ou(dat2, alpha = 0.5, sigmasq = 0.2,
                           root_value = 0, theta = theta)
head(expr)
```

```{r}
expr %>%
  left_join(dat2@data %>% select(node, status), by = "node") %>%
  filter(status == 1) %>%
  ggplot2::ggplot(ggplot2::aes(x = type, y = expr)) +
  ggplot2::geom_boxplot()
```

# Sequential (TypeWriter-style) barcodes

Sites are edited left-to-right, one at a time, as a Poisson process; an
independent whole-barcode silencing process competes against editing, and
dropout can additionally mask individual tip sites.

```{r}
k <- 5
chars <- c("A", "B", "C", "D")
lambda_vec <- c(0.3, 1.5)  # one rate per barcode copy

bc_seq <- typewriter_barcodes(dat2, k, lambda_vec,
                               silencing_rate = 0, dropout_p = 0,
                               m = length(lambda_vec), chars = chars)
head(bc_seq)
```

# Non-sequential (TiDeTree-style) barcodes

Each of the `k` sites evolves independently under a continuous-time Markov
chain, editing only within a bounded time window, with its own competing
silencing rate.

```{r}
k <- 10
edit_rates <- c(0.5, 0.3, 0.2)  # E = 3 possible edit outcomes
silencing_rate <- 0.05
edit_height <- 5    # window ends 5 time units before present
edit_duration <- 3  # window spans (edit_height - edit_duration, edit_height]

bc_nonseq <- sim_barcode_nonseq(dat2, k, edit_rates, silencing_rate,
                                 edit_height, edit_duration, dropout_p = 0)
head(bc_nonseq)
```

# Binary trait, with error and division-triggered spikes

A binary trait evolving under a 2-state CTMC, with optional
gamma-distributed mutation spikes at branches where a division occurred,
and optional false positive/negative error at the tips.

```{r}
bc_binary <- sim_binary_ontree(dat2, l = 10, gamma_spike = TRUE,
                                spike_mean = 0.3, spike_shape = 2,
                                alpha = 0.05, beta = 0.02)
head(bc_binary)
```
