---
title: "Simulating Lineage Barcodes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating Lineage Barcodes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 5,
  dpi = 300,
  out.width = "100%",
  eval = requireNamespace("ggtree", quietly = TRUE) && requireNamespace("ggplot2", quietly = TRUE)
)
```

```{r setup, message=FALSE}
library(scTreeSim)
library(ggtree)
library(ggplot2)
source("helpers.R")  # plotting helpers of the vignettes: plot_tree(), plot_barcodes() and a palette
```

Lineage barcodes are edited along the lineage tree of the cells and read out at its tips.
`scTreeSim` simulates them on a fixed tree, with three simulators:

| | `sim_barcode_seq()` | `sim_barcode_nonseq()` | `sim_barcode_generic()` |
|---|---|---|---|
| model | SciPhy | TiDeTree | both, and more |
| sites edited | one after the other, left to right | independently | either (`sequential` flag) |
| silencing | the whole barcode | each site separately | whole barcode or per site |
| editing window | no | yes (`edit_height`, `edit_duration`) | via time profiles |
| rates vary in time | no | window only | yes (`rate_window()`, ...) |
| rates depend on the cell type | no | no | yes (`by_type`) |
| dropout of a barcode at the tips | constant | constant | constant, or by type |

The simulators take the same kind of tree, and all return a table with one row per node. We use
a small multi-type tree (see `vignette("simulate_tree")`), so that the barcodes of all
cells are readable. The barcodes of the tips are shown as colour-coded heatmaps next to the tree (one row per
cell, one block per barcode, one column per site): a site is unedited (light grey), carries one of the
edit outcomes `1, 2, ...` (colours), or is missing (pink), i.e. silenced or dropped out. The simulators
return the same states as numbers (`0`, `1, 2, ...`) and a `missing_state` (here `"-"`) in tables, one row per node.

```{r tree}
origin_time <- 3.6
ntype <- 2
shape <- c(100, 1)
scale <- 1 / shape
r <- 0.11
Xi_as <- matrix(c(0.0, 0.0, r, 0.0), nrow = ntype, ncol = ntype)
Xi_s  <- matrix(c(1 - r, 0.0, 0.0, 1.0), nrow = ntype, ncol = ntype)

set.seed(23)
tree <- sim_adb_origin_samp(origin_time, scale, shape, death_prob = c(0.1, 0.1),
                            sampling_prob = 0.8, origin_type = 0,
                            asym_trans_prob = Xi_as, sym_trans_prob = Xi_s, collapse = FALSE)
stopifnot(!is.null(tree))  # a simulation with too few tips returns NULL: try another seed

plot_tree(tree, color = "type", show_nodes = TRUE)
```

The barcodes evolve along the branches, starting from an unedited barcode at the origin. The
tree needs to be ultrametric, i.e. all tips are sampled at the present, and `collapse = FALSE`
keeps the cell types on all branches (needed for rates that depend on the type, see below).

# Sequential editing: `sim_barcode_seq()`

In the SciPhy model the sites of a barcode are edited left to right, one at a time, at rate
`edit_rate`: only the first unedited site can be edited next, so edited sites always
form a prefix of the barcode. A barcode can also be silenced as a whole (`silencing_rate`), which
competes with editing, and from then on all its sites are missing in all descendants.

```{r seq}
edit_probs <- c(0.5, 0.3, 0.2)  # relative frequencies of the E = 3 edit outcomes
set.seed(2)
bc_seq <- sim_barcode_seq(tree, n_barcodes = 3, n_sites = 5,
                          edit_rate = 0.6, edit_probs = edit_probs,
                          silencing_rate = 0.05, dropout_prob = 0.1, missing_state = "-")
head(bc_seq, 3)
```

```{r seq-plot}
plot_barcodes(tree, bc_seq, n_edits = length(edit_probs))
```

Each cell has three barcodes. The edits always form a prefix of the barcode (the unedited
sites are at the right), cells that are close in the tree share the same edits, and a fully pink barcode was
silenced in an ancestor (shared by relatives) or dropped out at the tip (private to a cell).

# Independent sites in an editing window: `sim_barcode_nonseq()`

In the TiDeTree model every site is a separate target, edited independently of the others, and
editing is only possible in a window of time, as in systems where the editing machinery is switched on
for a limited period. The window opens `edit_height` time units before the present and lasts
`edit_duration`; each target is silenced independently at rate `silencing_rate`.

```{r nonseq}
set.seed(2)
bc_nonseq <- sim_barcode_nonseq(tree, n_targets = 10, edit_rate = 0.6, edit_probs = edit_probs,
                                edit_height = 3, edit_duration = 1.5,
                                silencing_rate = 0.05, dropout_prob = 0.1, missing_state = "-")
head(bc_nonseq, 3)
```

```{r nonseq-plot}
plot_barcodes(tree, bc_nonseq, n_edits = length(edit_probs))
```

Edits now appear at any site, and are shared by the cells that descend from the cell in which they
occurred. Because the window closes 1.5 time units before the present, the more recent divisions carry
no edits of their own: the barcodes of sister cells are identical, except for dropout, unlike with sequential editing
where editing continues until the present. This simulator returns one barcode of `n_targets` sites, and a silenced
site is missing on its own, not the whole barcode.

# One simulator for both: `sim_barcode_generic()`

`sim_barcode_generic()` has the arguments of both models, and switches between them with
`sequential`. With `sequential = TRUE` it is the model of `sim_barcode_seq()` (silencing and dropout
affect the whole barcode); with `sequential = FALSE` the sites are independent, as in
`sim_barcode_nonseq()`, but without a fixed window.

```{r generic}
set.seed(1)
bc_generic_seq <- sim_barcode_generic(tree, n_barcodes = 3, n_sites = 5,
                                      edit_rate = 0.6, edit_probs = edit_probs,
                                      silencing_rate = 0.05, dropout_prob = 0.1,
                                      sequential = TRUE)
set.seed(1)
bc_generic_ind <- sim_barcode_generic(tree, n_barcodes = 3, n_sites = 5,
                                      edit_rate = 0.6, edit_probs = edit_probs,
                                      silencing_rate = 0.05, dropout_prob = 0.1,
                                      sequential = FALSE)
```

```{r generic-plot, fig.height = 4.5}
plot_barcodes(tree, bc_generic_seq, n_edits = length(edit_probs)) +
  labs(title = "sequential = TRUE")
plot_barcodes(tree, bc_generic_ind, n_edits = length(edit_probs)) +
  labs(title = "sequential = FALSE")
```

The generic simulator is statistically equivalent to the specialised ones (its tests compare the
distributions of the barcodes of all three), and the specialised ones are faster. Everything
below is only available in the generic simulator.

The rates are given as numbers, or as `barcode_rate()` objects that add to a base rate a
**time profile** and/or **multipliers by cell type**. The rate of a branch of type `k` at time `t` since the
origin is `base * by_type[k + 1] * g(t)`.

# Rates that change in time

A time profile `g(t)` multiplies the base rate. Profiles are built from `rate_constant()`,
`rate_window()`, `rate_exp_decay()`, `rate_piecewise()` and `rate_fn()`, and can be combined with
`rate_product()`; see `?rate_profiles`. Time is measured since the origin of the tree.

```{r profiles, fig.height = 3}
e_start <- 1
e_end <- 2.5
profiles <- list(
  "window (1, 2.5)" = rate_window(e_start, e_end),
  "decay, half-life 1" = rate_exp_decay(1),
  "decay inside the window" = rate_product(rate_window(e_start, e_end), rate_exp_decay(1))
)
grid <- seq(0, origin_time, length.out = 400)
profile_df <- do.call(rbind, lapply(names(profiles), function(nm) {
  data.frame(profile = nm, time = grid, g = profiles[[nm]]$g(grid))
}))
ggplot(profile_df, aes(time, g)) +
  geom_line(color = type_cols[["0"]]) +
  facet_wrap(~ profile) +
  labs(x = "time since the origin", y = "multiplier g(t)") +
  theme_minimal()
```

An editing rate with the window profile only edits between times 1 and 2.5 since the origin, i.e. in
a window that is the same for all cells. Compare with a constant rate:

```{r window}
set.seed(1)
bc_window <- sim_barcode_generic(tree, n_barcodes = 1, n_sites = 10,
                                 edit_rate = barcode_rate(0.9, time = rate_window(e_start, e_end)),
                                 edit_probs = edit_probs, sequential = FALSE)
```

```{r window-plot}
plot_barcodes(tree, bc_window, n_edits = length(edit_probs)) +
  annotate("rect", xmin = -(origin_time - e_start), xmax = -(origin_time - e_end), ymin = -Inf, ymax = Inf,
           fill = type_cols[["2"]], alpha = 0.15)
```

Edits arise only in the shaded editing window (times 1 to 2.5 since the origin), so cells that split after
the window has closed have identical barcodes.
Silencing can have its own profile, e.g. `silencing_rate = barcode_rate(0.3, time = rate_exp_decay(1))`
for a process that fades out.

# Rates that depend on the cell type

`by_type` multiplies the rate by a number per cell type (`by_type[k + 1]` for type `k`). Here the cells of type 1 (pink) edit ten times faster than the cells of type 0 (blue), and
barcodes of type 1 cells drop out with probability 0.5. A dropout probability only applies at the tips, so it takes
`by_type` but no time profile.

```{r types}
set.seed(1)
bc_type <- sim_barcode_generic(tree, n_barcodes = 1, n_sites = 10,
                               edit_rate = barcode_rate(0.15, by_type = c(1, 10)),
                               edit_probs = edit_probs,
                               dropout_prob = barcode_rate(1, by_type = c(0, 0.5)),
                               missing_state = "-", sequential = FALSE)
```

```{r types-plot}
plot_barcodes(tree, bc_type, n_edits = length(edit_probs))
```

The cells of type 1 accumulate many more edits, and most of the missing (pink) sites are in the type 1
cells too. With independent sites, dropout masks the sites of a barcode one by one, whereas
for sequential editing it removes the whole barcode. The type of a branch is the type of the cell at its end,
so this needs a tree with `collapse = FALSE`; the root edge takes the type of the root.

# Choosing a simulator

* To simulate **SciPhy** barcodes, or the most simple barcodes with one editing rate: `sim_barcode_seq()`.
* To simulate **TiDeTree** barcodes with an editing window: `sim_barcode_nonseq()`.
* For **time-varying** (windows, decay) or **cell type-dependent** rates, or to switch between
  sequential and independent sites: `sim_barcode_generic()`.
